Formalizing Commutative Algebra in Coq: \\ Nakayama's Lemma\thanks{Source code for this work is available on the following site: \url{https://github.com/ku-sldg/algebra}}

### 2023-01-03 nakayama/paper.tex
Exposition edits throughout

Other familiar examples of commutative rings include the integers modulo a
fixed integer $n>0$, fields---commutative rings in which every nonzero element
has a multiplicative inverse---such as the rings of rational numbers, real
numbers, and complex numbers, and rings of polynomials in a variable $x$ with
integer coefficients, or with coefficients in a field.

@@

From this definition, one can see that a module over a field $F$ is precisely
an $F$-vector space, so the notion of a module over an arbitrary commutative
ring extends that of a vector space over a field. Finitely generated vector
spaces form the foundation for matrix algebra, and the extension of this notion
to module theory is needed to state Nakayama's Lemma. We call an $R$-module
$M$ \emph{finitely generated} if there exist a fixed finite number of elements
$u_1, \ldots, u_n$ of $M$ with the following property:  Given any  $w \in M$,
there exist $r_1, \ldots, r_n \in R$ for which  
\[w = r_1 u_1 + r_1 u_2 + \cdots + r_n u_n\text{.}\]
The set $\{u_1, \ldots, u_n\}$ is called a \emph{generating set} for the $M$ as
an $R$-module.

When $R = F$ is a field and $M = V$ is a finite-dimensional vector space over
$F$, one can choose $u_1, \ldots, u_n$ to be a basis for $V$, i.e.,
$n = \dim V$. In this case, the choice of scalar coefficients in the expression
above for $w \in V$ is unique. When $R$ is not a field, however, such an
expression is typically not unique.

@@

In order to state Nakayama's Lemma, we first explain some notation:  If $I$ is
an ideal of a commutative ring $R$ and $M$ is an $R$-module, then $IM$ is the
set of elements of the form $a_1 u_1 + a_2 u_2 + \cdots + a_k u_k$, where, for
some positive integer $k$, $a_1, \ldots, a_k \in I$ and
$u_1, \ldots, u_k \in M$. Notice that due to the absorption property of ideals,
$I M$ is an $R$-module contained in $M$.

If an $R$-module $M$ consists of only one element, this element must be its
additive identity $0$ as an abelian group under addition. The notation $M=0$
means that we are in this situation.

@@

When $R=F$ is a field, its unique maximal ideal is the zero ideal, and given
any vector space $M=V$ over $F$, the only linear combination of vectors with
coefficients in the zero ideal is the zero vector. Hence in this special case,
the hypothesis that  $M = \mathfrak{m} M$ is equivalent to the conclusion that
$M=0$. Hence Nakayama's Lemma describes one way that finitely generated modules
over commutative local rings are similar to vector spaces.

In general, the quotient $R/\mathfrak{m}$ of a local ring modulo its maximal
ideal $\mathfrak{m}$ is a field, and the quotient of a module $M$ modulo the
submodule $\mathfrak{m}M$ is an $R/\mathfrak{m}$-module, i.e.,
$M/\mathfrak{m}M$ is a vector space over $R/\mathfrak{m}$. Nakayama's Lemma
implies that if $M$ is finitely generated, then bases for $M/\mathfrak{m}M$
corresponds, via lifting, to minimal sets of generators of $M$.

@@

\section{Formalization}
We start by describing our process of formalizing the required algebraic
structures detailed in the previous section. Then, with that in hand, we move
on to the formal proof of Nakayama's Lemma.

@@

Our foundation begins by defining a semigroup class, which declares a binary
operation to be associative. From here, we build up through monoids, which
introduce identities, to groups, which introduce inverses. Note the double
equals ``$==$'' appearing in these definitions is notation for an arbitrary
equivalence relation over the group's carrier set, which acts as equality.

\begin{minted}{coq}
  Infix "==" := equiv (at level 60, no associativity).
  Class Semigroup := {
    semigroup_assoc:
      forall (a b c: Carrier),
        a <o> b <o> c == a <o> (b <o> c);
  }.
  Class Monoid := {
    monoid_semigroup :> Semigroup equiv op;
    monoid_ident_l:
      forall (a: Carrier), ident <o> a == a;
    monoid_ident_r:
      forall (a: Carrier), a <o> ident == a;
  }.
  Class Group := {
    group_monoid :> Monoid equiv op ident;
    group_inv_l:
      forall (a: Carrier), inv a <o> a == ident;
    group_inv_r:
      forall (a: Carrier), a <o> inv a == ident;
  }.
\end{minted}
Lines such as  ``\verb|monoid_semigroup :> Semigroup equiv op;|'' simply coerce
the monoid typeclass into a semigroup.

%While we found later on that we did not need quotients, 
While in the end, our formal proof does not require calling upon quotients of
algebraic structures, quotient rings and quotient modules are fundamental to
commutative algebra, and one can use them to construct alternate proofs of
Nakayama's Lemma. It is worth pointing out that we have formalized quotients of
algebraic objects in Coq using typeclasses, which appear to work rather nicely.

An algebraic quotient is, roughly, the set of equivalence classes of an
algebraic structure with respect to an equivalence relation on its elements,
for which the set of equivalence classes inherit the same kind of algebraic
structure. For example, consider the quotient of a group modulo a subgroup,
i.e., a subset of elements of the group that it itself a group under the group
operations. Under equivalence relation on the group, every element of the
subgroup must be in the same equivalence class as the identity. With \texttt{P}
the predicate for the subgroup, there are two ways to make an equivalence
relation from this description.

@@

When these two relations coincide, then we can prove that this common
equivalence relation actually preserves the group structure. Subgroups which
have this property are called \emph{normal subgroups}.

@@

The importance of quotients in commutative algebra motivates our use of
equivalence relations to define the components of a group structure. If one
were to instead use the regular Leibniz equality, it would be difficult to
identify a quotient group and another group. However, by  defining a group in
terms of an arbitrary equivalence relation, in our theory a quotient group is 
simply defined as a group, but under an equivalence relation that is not the
usual equality. Not much is lost due to Coq's rewrite tactics for
setoids--types equipped with an equivalence relation---can still be called
upon.

Moving onward, rings form the next step in our algebraic hierarchy; a ring has
two binary operations: addition, which must be commutative, and multiplication,
which need not be commutative in general. In our formulation, rings must have a
multiplicative identity. Next, we formalized the definition of a commutative
ring, further requiring commutativity of multiplication. At this point, we
formalized the notion of an ideal of a commutative ring, a subgroup of the ring
under addition that satisfies the absorption property under multiplication,
i.e., \(r a\) is in the ideal for every element \(a\) of the ideal, and every
element \(r\) of the commutative ring. We also used this to formalize the
notion of a quotient ring $R/I$, where $I$ is an ideal of a commutative ring
$R$.

@@

We then were able to take advantage of the definition of a maximal ideal
to formally define a local
ring, i.e., a commutative ring with a single maximal ideal.

@@

At this point we formalized the definition of a module over a commutative ring,
the commutative-algebraic generalization of the notion of a vector space over a
field. Nakayama's Lemma is a statement about finitely generated modules, and
hence we must formalize the notion of a scalar combination of a finite
collection of elements, $u_1, \ldots, u_n$, of an $R$-module $M$, i.e.,
expressions of the form $r_1 u_1 + r_1 u_2 + \cdots + r_n u_n$, where each
$r_i \in R$.

@@

Next, given an ideal \(I\) of a commutative ring $R$ and and an $R$-module
\(M\),  we defined the submodule $IM$, the set consisting of all scalar
combinations of elements of \(M\) whose coefficients are in  \(I\). We
represented this in Coq as a predicate over \(M\).

@@

The use of  ``\verb|Forall P coeffs|'' ensures that every element of the
coefficient list \texttt{coeffs} satisfies the predicate \texttt{P}.

@@

We formally proved that an ideal $I$ of a commutative ring $R$ that contains a
unit $x$, then $I$ must be the trivial ideal, i.e., the entire ring.  The
informal logic is as follows:  By the absorption property, since $x \in I$, we
have that $x \cdot x^{-1} = 1 \in I$.  Hence for every element $r$ of $R$,
$r = r \cdot 1$ is also in $I$, i.e., $I=R$.

Every non-unit element of a commutative ring is contained in some maximal
ideal. This fact relies on the Axiom of Choice. The following standard informal
proof calls upon Zorn's lemma, which is equivalent to the Axiom of Choice, and
says that given a partially ordered set $S$, if every chain in $S$ has an upper
bound, then $S$ must have at least one maximal element.

@@

This argument has potentially infinitely many steps. A proof with infinitely
many steps is a problem in Coq. To see this from the above argument, suppose we
wrote a proof tactic \mintinline{coq}{generate_larger_ideal} that allows us to
go from one non-maximal ideal \(I_{k}\) which contains \(x\) to another ideal
\(I_{k+1}\) which also contains \(x\). Provided our first ideal \(I_{1}\) is
not maximal, our first invocation of \mintinline{coq}{generate_larger_ideal}
will produce another ideal \(I_{2}\). Our new ideal \(I_{2}\) may also not be
maximal, requiring that we invoke \mintinline{coq}{generate_larger_ideal}
again. This can lead to an infinite number of calls to the tactic
\mintinline{coq}{generate_larger_ideal} which will occupy any machine's
processor(s) forever, and further, the infinite number of ideals in the proof
state will fill any machine's memory. After we have overwhelmed any system with
these infinite number of calls to a single proof tactic, we still have to
invoke more tactics to finish the proof with an application of Zorn's lemma,
which finally guarantees us a maximal ideal containing \(x\). To summarize, we
are forced to run the following proof script which will never get past calling
\mintinline{coq}{generate_larger_ideal} infinitely many times.
\begin{minted}{coq}
  repeat generate_larger_ideal. (* must be called infinitely often *)
  zorns_lemma. (* called after the above non-terminating call finishes *)
\end{minted}
We chose to avoid this infinite proof script issue by including an axiom that
any non-unit \(x\) of a ring must be contained in some maximal ideal.

>>>>

Other familiar examples of commutative rings include the integers modulo a
fixed integer $n>0$, fields---commutative rings in which every nonzero element
has a multiplicative inverse---such as the rings of rational numbers $\mathbb{Q}$, real
numbers $\mathbb{R}$, and complex numbers $\mathbb{C}$, and the rings
 of polynomials in a variable $x$ with
integer coefficients, or with coefficients in a field.

A \emph{subgroup} of a group $G$ is a subset  $H$ of $G$ that is itself a 
group when the binary operation of $G$ is restricted to $H$.
Similarly, a \emph{subring} of a ring $R$
is a subset $S$ of $R$ that forms a group when the operations on $R$ are 
restricted to $S$, using the same multiplicative identity.  
One has the sequence of subrings $\mathbb{Z} \subseteq \mathbb{Q} \subseteq \mathbb{R} \subseteq \mathbb{C}$.

@@

From this definition, one can see that a module over a field $F$ is precisely
an $F$-vector space, so the notion of a module over an arbitrary commutative
ring extends that of a vector space over a field. 
An $R$-\emph{submodule} $N$ of a module $M$ is simply a subgroup of $M$ that inherits 
the scalar multiplication of $N$, i.e., $r \cdot w$ is an element of $N$ whenever $w \in N$.
Therefore, returning to the case that $F$ is a field,
the $F$-submodules of a vector space over $F$ are precisely its vector subspaces.

Finitely generated vector spaces form the foundation for matrix algebra, and the extension of this notion
to module theory is needed to state Nakayama's Lemma. We call an $R$-module
$M$ \emph{finitely generated} if there exist a fixed finite list of elements
$u_1, \ldots, u_n$ of $M$ such that every element of $M$ is a scalar combination 
of the $u_i$.  In other words, 
given any  $w \in M$,
there exist $r_1, \ldots, r_n \in R$ for which  
\[w = r_1 u_1 + r_1 u_2 + \cdots + r_n u_n\text{.}\]
The set $\{u_1, \ldots, u_n\}$ is called a \emph{generating set} for the $M$ as
an $R$-module.

It is straightforward to see that a finitely generated module over a field is  
simply a finite dimensional vector space.
Moreover, if $V$ is a finite dimensional vector space over a field 
$F$, one can choose the generators $u_1, \ldots, u_n$ to be a basis for $V$, so that  
$n = \dim V$. In this case, the choice of scalar coefficients in the expression
above for $w \in V$ is unique. When $R$ is not a field, however, such an
expression is typically not unique.

@@

In order to informally state Nakayama's Lemma, we first explain some of notational 
conventions.  In this dicussion, suppose that  $I$ is an ideal of a commutative ring $R$, 
and that $M$ is an $R$-module. Then $IM$ denotes the set of all scalar combinations of 
elements in $M$ with coefficients in $I$.  That is, $IM$ is the set of  elements of the form, 
for some positive integer $k$,
 \[a_1 u_1 + a_2 u_2 + \cdots + a_k u_k\] 
where $a_1, \ldots, a_k \in I$ and
$u_1, \ldots, u_k \in M$. Observe the absorption property of ideals implies that 
$I M$ is an $R$-submodule of $M$.

If an $R$-module $M$ consists of just one element, this element must be its
additive identity $0$, by virtue of the fact that $M$ is an abelian group under addition. 
The notation $M=0$ is used to indicate that we are in this situation.

@@

When $R=F$ is a field, its unique maximal ideal is the zero ideal.  
Given any vector space $M=V$ over $F$, the only linear combination of vectors with
coefficients in the zero ideal is the zero vector. Hence in this case, regardless of the choice of 
vector space, the hypothesis that  $M = \mathfrak{m} M$ is equivalent to the conclusion that
$M=0$. 
In particular, the hypothesis that $V$ is finitely generated, which is equivalent to the
assumption that it is a finite-dimensional $F$-vector space, is not required in this setting. 
Therefore, Nakayama's Lemma describes one way that finitely generated modules
over commutative local rings are similar to vector spaces.

The quotient $R/\mathfrak{m}$ of a commutative ring $R$ modulo any maximal ideal 
$\mathfrak{m}$ is a field; in fact, this property characterizes maximal ideals. 
Moreover, given an arbitrary $R$-module $M$, its quotient modulo the
submodule $\mathfrak{m}M$ is an $R/\mathfrak{m}$-module, i.e.,
$M/\mathfrak{m}M$ is a vector space over $R/\mathfrak{m}$. Nakayama's Lemma
implies that if $M$ is finitely generated, then bases for $M/\mathfrak{m}M$
corresponds, via lifting, to minimal sets of generators of $M$.

@@

\section{Formalization}
We start by describing our process of formalizing the necessary algebraic
structures detailed in the previous section. Then, with that in hand, we move
on to the formal proof of Nakayama's Lemma.

@@

\paragraph{Semigroups.} Our foundation begins by defining a semigroup class, which declares a binary
operation to be associative. From here, we build up through monoids, which
introduce identities, to groups, which introduce inverses. Note the double
equals ``$==$'' appearing in these definitions is notation for an arbitrary
equivalence relation over the group's carrier set, which acts as equality.

\begin{minted}{coq}
  Infix "==" := equiv (at level 60, no associativity).
  Class Semigroup := {
    semigroup_assoc:
      forall (a b c: Carrier),
        a <o> b <o> c == a <o> (b <o> c);
  }.
  Class Monoid := {
    monoid_semigroup :> Semigroup equiv op;
    monoid_ident_l:
      forall (a: Carrier), ident <o> a == a;
    monoid_ident_r:
      forall (a: Carrier), a <o> ident == a;
  }.
  Class Group := {
    group_monoid :> Monoid equiv op ident;
    group_inv_l:
      forall (a: Carrier), inv a <o> a == ident;
    group_inv_r:
      forall (a: Carrier), a <o> inv a == ident;
  }.
\end{minted}
The lines such as  ``\verb|monoid_semigroup :> Semigroup equiv op;|'' simply coerce
the monoid typeclass into a semigroup.

\paragraph{Algebraic quotients.}
While in the end, our formal proof does not call upon quotients of
algebraic structures, alternative proofs of Nakamaya's Lemma take advantage of 
 the structures quotient rings and quotient modules.  As algebraic 
 quotient structures are also fundamental to commutative algebra,
it is work pointing out that we have formalized quotients of groups, rings, and modules 
  using typeclasses, which appear to work rather smoothly.

An algebraic quotient is, roughly, the set of equivalence classes of an
algebraic structure with respect to an equivalence relation on its elements,
for which the set of equivalence classes inherit the same kind of algebraic
structure. 
As an example, consider the quotient of a group modulo a subgroup,
i.e., a subset of elements of the group that it itself a group using the same operations. 
Under this equivalence relation on the group, every element of the
subgroup must be in the same equivalence class as the identity. 
For instance, after taking the quotient of the ring of integers $\mathbb{Z}$ by the
subgroup $n \mathbb{Z}$ consisting of all multiples of an integer $n$, one obtains the
group of integers modulo $n$, often denoted $\mathbb{Z}/n\mathbb{Z}$.
In general, with \texttt{P}
the predicate for the subgroup, there are two ways to make an equivalence
relation from this description.

@@

When these two equivalence relations coincide, then we can prove that this common
 relation actually preserves the group structure. Subgroups of a group that satisfy 
this property are called \emph{normal subgroups}.

@@

The utility--and, as a consequence, the ubiquity--of quotients in commutative algebra 
motivates our choice to use equivalence relations to define the components of a group structure. If one
were to instead use the traditional Leibniz equality, it would be difficult to
identify a quotient group with another group. However, by  defining a group in
terms of an arbitrary equivalence relation, in our theory a quotient group is 
simply defined as a group, but under an equivalence relation that is not the
usual equality. Not much is lost in adopting this convention thanks to Coq's 
rewrite tactics for setoids--types equipped with an equivalence relation---can 
still be called
upon.

\paragraph{Rings and their ideals.}
Moving onward, rings form the next step in our algebraic hierarchy; a ring has
two binary operations: addition, which must be commutative, and multiplication,
which need not be commutative in general. In our formulation, rings must have a
multiplicative identity. Next, we formalized the definition of a commutative
ring, further requiring commutativity of multiplication. At this point, we
formalized the notion of an ideal of a commutative ring, a subgroup of the ring
under addition that satisfies the absorption property under multiplication,
i.e., \(r a\) sits in the ideal whenever \(a\) of the ideal, regardless of the choice of 
element \(r\) of the commutative ring. We also used this to formalize the
notion of a quotient ring $R/I$, where $I$ is an ideal of a commutative ring
$R$.

@@

\noindent Next, we were able to employ this definition of a maximal ideal
to formally define a local
ring, i.e., a commutative ring with a single maximal ideal.

@@

\paragraph{Modules over rings.}
At this point, we formalized the definition of a module over a commutative ring,
the commutative-algebraic generalization of the notion of a vector space over a
field. Nakayama's Lemma is a statement about finitely generated modules, and
hence we formalized the notion of a scalar combination of a finite
collection of elements, $u_1, \ldots, u_n$ a module $M$ over a commutative ring $R$, i.e.,
expressions of the form $r_1 u_1 + r_1 u_2 + \cdots + r_n u_n$, where each
$r_i$ is an element of $R$.

@@

Finally, we defined the submodule $IM$ of a module $M$ over a commutative ring $R$, 
where $I$ is an arbitrary ideal of $R$.
Recall that when $R$ is local and $I$ is its unique maximal ideal, this submodule 
appears in the statement of Nakayama's Lemma. 
In general, $IM$ is the set consisting of all scalar
combinations of elements of \(M\) whose coefficients are in  \(I\). We
represent this in Coq as a predicate over \(M\).

@@

\noindent The use of  ``\verb|Forall P coeffs|'' ensures that every element of the
coefficient list \texttt{coeffs} satisfies the predicate \texttt{P}.

@@

We formally prove that an ideal $I$ of a commutative ring $R$ that contains a
unit $x$, then $I$ must be the trivial ideal, i.e., the entire ring.  The
informal logic is as follows:  Due to the absorption property of $I$,  
$x \cdot x^{-1} = 1 \in I$.  Hence for every element $r$ of $R$,
$r = r \cdot 1$ is also in $I$, i.e., $I=R$.

Every non-unit element of a commutative ring is contained in some maximal
ideal, a fact that relies on the Axiom of Choice.

The following standard informal
proof calls upon Zorn's lemma, which is equivalent to the Axiom of Choice, and
says that given a partially ordered set $S$, if every chain in $S$ has an upper
bound, then $S$ must have at least one maximal element.

@@

The above argument has potentially infinitely many steps. A proof with infinitely
many steps is a problem in Coq. To see this from the above argument, suppose we
wrote a proof tactic \mintinline{coq}{generate_larger_ideal} that allows us to
go from one non-maximal ideal \(I_{k}\) which contains \(x\) to another ideal
\(I_{k+1}\) which also contains \(x\). Provided our first ideal \(I_{1}\) is
not maximal, our first invocation of \mintinline{coq}{generate_larger_ideal}
will produce another ideal \(I_{2}\). Our new ideal \(I_{2}\) may also not be
maximal, requiring that we invoke \mintinline{coq}{generate_larger_ideal}
again. This can lead to an infinite number of calls to the tactic
\mintinline{coq}{generate_larger_ideal} which will occupy any machine's
processor(s) forever, and further, the infinite number of ideals in the proof
state will fill any machine's memory. After we have overwhelmed any system with
these infinite number of calls to a single proof tactic, we still have to
invoke more tactics to finish the proof with an application of Zorn's lemma,
which finally guarantees us a maximal ideal containing \(x\). To summarize, we
are forced to run the following proof script which will never get past calling
\mintinline{coq}{generate_larger_ideal} infinitely many times.
\begin{minted}{coq}
  repeat generate_larger_ideal. (* must be called infinitely often *)
  zorns_lemma. (* called after the above non-terminating call finishes *)
\end{minted}
We chose to avoid this infinite proof script issue by including an axiom that
any non-unit \(x\) of a ring must be contained in some maximal ideal.
